Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-12-08 paper/main.tex
added histogram of "gaps within sight"
moved comment about no gap near GC

@@

  The simulations presented in this paper suggest that, in the last 5 Gyr of evolution, Palomar~5's stream could have experienced multiple close encounters with other Galactic globular clusters, some of which are able to create gaps -- even a few kpc wide -- in its tails.

  Unfortunately, none of the predicted gaps is located in regions of the tails that have been observed, and this is especially true for the leading tail, which result shorter than the trailing tail in the observations \citep[see][for a possible explanation of this feature]{2017NatAs...1..633P}. Currently, the presence -- or not -- of gaps in the observed portion of Pal~5 tails is debated: \citet{2016ApJ...819....1I} found no statistically significant gap in Pal~5 tails, while \citet{2017MNRAS.470...60E}, analyzing the same data-set as \citet{2016ApJ...819....1I}, suggested the presence of a few dips and gaps in the tails, at angular distances between $2^\circ$ and $9^\circ$ from the cluster center \citep[see also][]{ 2020ApJ...889...70B}. We cannot reproduce any of these features with clusters flybys in our simulations, since all our gaps are found at much larger angular and spatial distance from the cluster itself. There may be a few reasons for the absence of gaps in portion of the tails closer to the cluster center. First, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuthal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occurred close to the cluster were inconsequential for gap formation. Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Pal~5, the tails are very thin, with a typical thickness being less than 100 pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Pal~5, 5 Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5 Gyr of evolution, our N-body system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent.  This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5 tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

>>>>

  \subsection{Comparison to observations}

    We briefly compare our simulated gaps to the literature on Paloamr~5. In the bottom panel of Fig.~\ref{fig:stream_on_sky}, we compare the tracks of Palomar~5 that were compiled in Galstreams by Fig.~\ref{2023MNRAS.520.5225M}. The mid-point positions of the streams do not have the same galacto-centric positions as Palomar~5 from  Baumgardt's catalog and are thus offset when projected into tail coordinates. Nonetheless, we note that the lengths of the streams are at max 6~kpc on either side of the globular cluster. We apply a threshold of 5.75~kpc on either side to consider the gaps that could occur in currently observed regions of Paloamr~5. This is presented in Fig.~\ref{fig:GapsWithinSight}.
      \begin{figure}
        \centering
        \includegraphics[width=\linewidth]{GapWithinSight.png}
        \caption{The distribution of gaps in absolute distance from the center of mass of Palomar~5's globular cluster. A threshold value of $\sim$~5.75~kpc was selected as distance threshold that corresponds to the current observed length. Distances beyond this are shaded. }
        \label{fig:GapsWithinSight}
      \end{figure}

    We observe that there are no gaps near within 3~kpc to the cluster within our simulation. There may be a few reasons for the absence of gaps in portion of the tails closer to the cluster center. First, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuthal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occurred close to the cluster were inconsequential for gap formation.

    Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Pal~5, the tails are very thin, with a typical thickness being less than 100 pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Pal~5, 5 Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5 Gyr of evolution, our system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent. This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5 tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

@@

  The simulations presented in this paper suggest that, in the last 5~Gyr of evolution, Palomar~5's stream could have experienced multiple close encounters with other Galactic globular clusters, some of which are able to create gaps -- even a few kpc wide -- in its tails.

  Unfortunately, none of the predicted gaps is located in regions of the tails that have been observed, and this is especially true for the leading tail, which result shorter than the trailing tail in the observations \citep[see][for a possible explanation of this feature]{2017NatAs...1..633P}. \textbf{NOT TRUE. Good thing I checked.}

  Currently, the presence -- or not -- of gaps in the observed portion of Pal~5 tails is debated: \citet{2016ApJ...819....1I} found no statistically significant gap in Pal~5 tails, while \citet{2017MNRAS.470...60E}, analyzing the same data-set as \citet{2016ApJ...819....1I}, suggested the presence of a few dips and gaps in the tails, at angular distances between $2^\circ$ and $9^\circ$ from the cluster center \citep[see also][]{ 2020ApJ...889...70B}.

  We cannot reproduce any of these features with clusters flybys in our simulations, since all our gaps are found at much larger angular and spatial distance from the cluster itself.
